Long-form recording of infant body position in the home using wearable inertial sensors

### 2023-03-24 manuscript.Rmd
fixed timeline data inclusion

```{r timelines, warning=FALSE, fig.cap="Full-day timelines for each individual recording session, split by (A) younger infants and (B) older infants. Each participant's timeline shows a stacked bar graph with the proportion of time spent in each of the body position for every 5-minute period throughout the day, based on group model predictions of body position. The x-axis shows time of day. Caregiver-reported naps are marked by gray bars; blank gaps indicate caregiver-reported times that the sensor garment was removed for diaper changes, baths, or trips out of the house.", fig.align="center", fig.width=7.75, fig.height=8, cache=TRUE, cache.extra=tools::md5sum("data/imu-compiled.csv")}
timeline_data <- read_csv("data/imu-compiled.csv") %>% 
  mutate(id_uni = factor(id*100+session),
         group = factor(ifelse(age <= 8, "Younger", "Older")),
         nap_time = ifelse(is.na(sit_time), nap_time, NA)) %>% 
  filter(id_uni != 10401) %>% 
  pivot_longer(cols = nap_time:upright_time, names_to = "position", values_to = "prop") %>% 
  mutate(Position = factor(position,
                           levels = c("nap_time","upright_time", "sit_time", "prone_time", "supine_time", "held_time"),
                           labels = c("Nap","Upright", "Sitting", "Prone", "Supine", "Held"))) %>% 
  arrange(Position, age)

@@

Among participants who wore the sensors from morning to bedtime, the majority of the time during the day either resulted in usable body position data or was excluded due to caregiver-reported naps (gray periods in Figure \@ref(fig:timelines)). The total length of the recording period ranged from `r apa_num(min(coverage$total))`-`r apa_num(max(coverage$total))` hours with a median of `r apa_num(median(coverage$total))` hours. Nap times reported during the recording period ranged from `r apa_num(min(coverage$naptime))`-`r apa_num(max(coverage$naptime))` hours with a median of `r apa_num(median(coverage$naptime))` hours. Body position data was available to describe `r apa_num(min(coverage$prop_accounted)*100, digits = 1)`-`r apa_num(max(coverage$prop_accounted)*100, digits = 1)`% of the awake portion of the recording period (median = `r apa_num(median(coverage$prop_accounted)*100, digits = 1)`%). Younger infant \#13 had the least portion of the waking day accounted for; the family left the house to run errands during the majority of the recording period, and the infant napped during much of the remaining time at home. Overall, the recordings produced a median of `r apa_num(median(coverage$hours_recorded))` hours of motion data, with the entire dataset totaling `r apa_num(sum(coverage$hours_recorded))` hours. Even the shortest full recording (`r apa_num(min(coverage$hours_recorded))` hours) exceeded the longest observational sessions in past work in which an experimenter operated a camera.

@@

```{r age, warning=FALSE, fig.cap="Age differences in daily body position predicted from (A) group models and (B) individual models. Each circle represents one full-day recording session's proportion of time in each body position (y-axis) against age in months (x-axis). Black lines indicate best fit regression lines indicating decreases in supine time and increases in sitting and upright time with age.", fig.align="center", fig.width=7.5, fig.height=7, cache=TRUE, cache.extra=tools::md5sum("data/imu-compiled.csv")}
fullday_group <- read_csv("data/imu-compiled.csv") %>% 
  mutate(id_uni = factor(id*100+session),
         group = factor(ifelse(age <= 8, "Younger", "Older")),
         nap_time = ifelse(is.na(sit_time), nap_time, NA)) %>% 
  filter(id_uni != 10401)
fullday_split <- read_csv("data/imu-compiled-split.csv") %>% 
  mutate(id_uni = factor(id*100+session),
         group = factor(ifelse(age <= 8, "Younger", "Older")),
         nap_time = ifelse(is.na(sit_time), nap_time, NA)) %>% 
  filter(id_uni != 10401)
fullday <- list("group" = fullday_group, "split" = fullday_split)

>>>>

```{r timelines, warning=FALSE, fig.cap="Full-day timelines for each individual recording session, split by (A) younger infants and (B) older infants. Each participant's timeline shows a stacked bar graph with the proportion of time spent in each of the body position for every 5-minute period throughout the day, based on group model predictions of body position. The x-axis shows time of day. Caregiver-reported naps are marked by gray bars; blank gaps indicate caregiver-reported times that the sensor garment was removed for diaper changes, baths, or trips out of the house.", fig.align="center", fig.width=7.75, fig.height=8, cache=TRUE, cache.extra=tools::md5sum("data/imu-compiled.csv")}
timeline_data <- read_csv("data/imu-compiled.csv") %>% 
  mutate(id_uni = factor(id*100+session),
         group = factor(ifelse(age <= 8, "Younger", "Older")),
         nap_time = ifelse(is.na(sit_time), nap_time, NA)) %>% 
  pivot_longer(cols = nap_time:upright_time, names_to = "position", values_to = "prop") %>% 
  mutate(Position = factor(position,
                           levels = c("nap_time","upright_time", "sit_time", "prone_time", "supine_time", "held_time"),
                           labels = c("Nap","Upright", "Sitting", "Prone", "Supine", "Held"))) %>% 
  arrange(Position, age)

@@

Among participants who wore the sensors from morning to bedtime, the majority of the time during the day either resulted in usable body position data or was excluded due to caregiver-reported naps (gray periods in Figure \@ref(fig:timelines)). The total length of the recording period ranged from `r apa_num(min(coverage$total))`-`r apa_num(max(coverage$total))` hours with a median of `r apa_num(median(coverage$total))` hours. Nap times reported during the recording period ranged from `r apa_num(min(coverage$naptime))`-`r apa_num(max(coverage$naptime))` hours with a median of `r apa_num(median(coverage$naptime))` hours. Body position data was available to describe `r apa_num(min(coverage$prop_accounted)*100, digits = 1)`-`r apa_num(max(coverage$prop_accounted)*100, digits = 1)`% of the awake portion of the recording period (median = `r apa_num(median(coverage$prop_accounted)*100, digits = 1)`%). Younger infant \#14 had the least portion of the waking day accounted for; the family left the house to run errands during the majority of the recording period, and the infant napped during much of the remaining time at home. Overall, the recordings produced a median of `r apa_num(median(coverage$hours_recorded))` hours of motion data, with the entire dataset totaling `r apa_num(sum(coverage$hours_recorded))` hours. Even the shortest full recording (`r apa_num(min(coverage$hours_recorded))` hours) exceeded the longest observational sessions in past work in which an experimenter operated a camera.

@@

```{r age, warning=FALSE, fig.cap="Age differences in daily body position predicted from (A) group models and (B) individual models. Each circle represents one full-day recording session's proportion of time in each body position (y-axis) against age in months (x-axis). Black lines indicate best fit regression lines indicating decreases in supine time and increases in sitting and upright time with age.", fig.align="center", fig.width=7.5, fig.height=7, cache=TRUE, cache.extra=tools::md5sum("data/imu-compiled.csv")}
fullday_group <- read_csv("data/imu-compiled.csv") %>% 
  mutate(id_uni = factor(id*100+session),
         group = factor(ifelse(age <= 8, "Younger", "Older")),
         nap_time = ifelse(is.na(sit_time), nap_time, NA))

fullday_split <- read_csv("data/imu-compiled-split.csv") %>% 
  mutate(id_uni = factor(id*100+session),
         group = factor(ifelse(age <= 8, "Younger", "Older")),
         nap_time = ifelse(is.na(sit_time), nap_time, NA))

fullday <- list("group" = fullday_group, "split" = fullday_split)

### 2023-03-24 manuscript.tex
fixed timeline data inclusion

Most participants were tested in a single session (\emph{n} = 16), but 6 participants contributed between 2-4 sessions as part of an ongoing longitudinal study. Only 1 session was excluded due to a technical error---one of the four IMU sensors failed to record, resulting in an unusable set of data for classification. Across the two age groups, we report data on a total of 33 sessions, with 13 sessions from younger infants and 20 sessions from older infants. Across sessions, younger infants' age ranged from 3.8 to 7.2 months (\emph{M} = 5.0) and older infants' age ranged from 10.7 to 14.2 months (\emph{M} = 11.7) .

@@

We report three sets of results based on 33 full-day testing sessions resulting in a total of 294 hours of movement recording. First, we focus on the accuracy of group and individual models trained and validated from data during the proximal comparison period. This period is unique in having the highest degree of temporal synchrony between video and motion data, allowing us to assess the accuracy of individual body position events. Second, we applied models trained from data during the proximal comparison to predict infant behavior during the distal comparison, when coarser synchrony between video and motion data was available. This distal comparison is critical because it provides the first-ever test of how well infant body classification models predict behavior beyond the initial, supervised period. Data from the first two sets of analyses confirm that body position classification models are accurate across a long period of time, suggesting that data from the entire day will be valid. In the third set of results we examine the data quality of full-day recordings and test whether full-day recordings capture well-established age differences in body position experiences.

@@

Whereas the proximal analyses used all 33 sessions, this was not possible in the distal comparison. Because the start of the visit was scheduled during a time the infant was awake, it was common for a nap to be needed following the proximal period. Nine sessions were excluded because the infant was either napping or otherwise not on camera during the entire 90-minute distal recording. Three additional sessions were excluded because a caregiver accidentally turned off the video camera (\emph{n} = 1) or (purposefully) left the house (\emph{n} = 2). This left 22 sessions with usable distal comparison data.

@@

Figure \ref{fig:timelines} depicts body position timelines across the day for each session, divided into younger (A) and older infants (B). Predictions from the group model were used to ensure that motion data were classified consistently across all sessions. Session start times ranged from 08:55 to 13:20 with a median of 10:20. Although the equipment was always dropped off in the morning, sometimes infants were unexpectedly asleep, leading to a few sessions in which infants began wearing the sensors later than intended (such as \#11 in the younger group). With two exceptions, recordings lasted until the infants' bed time. Older participant \#1 had the equipment picked up on the same day rather than the next day, so the recording ended at 17:00. Older participant \#11 wore the equipment in the morning, but the family left the house at 10:15 and remained out for the rest of the day, choosing not to put the equipment back on when they returned home in the late afternoon. Among those participants who wore the equipment until bedtime, recordings ended between 17:25 and 22:15 with a median of 19:20.

Among participants who wore the sensors from morning to bedtime, the majority of the time during the day either resulted in usable body position data or was excluded due to caregiver-reported naps (gray periods in Figure \ref{fig:timelines}). The total length of the recording period ranged from 6.92-11.75 hours with a median of 9.08 hours. Nap times reported during the recording period ranged from 0.00-4.99 hours with a median of 2.22 hours. Body position data was available to describe 18.3-100.0\% of the awake portion of the recording period (median = 100.0\%). Younger infant \#13 had the least portion of the waking day accounted for; the family left the house to run errands during the majority of the recording period, and the infant napped during much of the remaining time at home. Overall, the recordings produced a median of 6.15 hours of motion data, with the entire dataset totaling 201.06 hours. Even the shortest full recording (3.06 hours) exceeded the longest observational sessions in past work in which an experimenter operated a camera.

Considering that body position annotation takes 2-5 hours of labor per 1 hour of behavior, the 201 hours of data we recorded would have taken 402-1005 hours of labor to annotate from video. However, only an estimated 62-155 hours of labor was needed to annotate the initial hour of each session's video that was used to train the machine learning models---an immense savings in the human cost of annotation that allows the method to scale to larger sample sizes and recording durations.

@@

\begin{tabular}{lllll}
\toprule
 & \multicolumn{2}{c}{Group} & \multicolumn{2}{c}{Individual} \\
\cmidrule(r){2-3} \cmidrule(r){4-5}
Position & Younger & Older & Younger & Older\\
\midrule
Supine & 38.5\% (24.0) & 14.0\% (8.4) & 40.2\% (31.6) & 10.0\% (9.5)\\
Prone & 14.5\% (13.8) & 14.4\% (6.0) & 12.5\% (10.2) & 16.9\% (10.6)\\
Sitting & 25.4\% (12.4) & 44.4\% (10.1) & 18.8\% (16.3) & 46.9\% (13.3)\\
Upright & 7.7\% (9.3) & 18.6\% (7.4) & 10.4\% (13.6) & 18.7\% (8.4)\\
Held & 13.1\% (7.4) & 8.5\% (5.4) & 17.1\% (20.8) & 7.4\% (7.6)\\
\bottomrule
\end{tabular}

@@

The most straightforward comparisons between the two investigations are upright and prone, because they were defined identically. In Franchak (2019), upright time was 0.6-5.5\% of the time for infants 3-6 months, increasing to 22.0\% at 12 months. Results from the current study were similar: younger infants' sitting time estimated from full-day group models was 7.72\% compared with 18.57\% for the older group. In Franchak (2019), prone time was 2.9-9.2\% of the time for infants 3-6 months, and was 7.2\% at 12 months. In the current study, prone time was 14.46\% for younger infants and 14.45\% for older infants. Held, sitting, and supine are more difficult to compare because prior work (Franchak, 2019) considered held as any time that infants were in caregivers' arms or on their laps, however, we counted laying or sitting on caregivers' laps as supine and sitting, respectively. Further complicating the comparison, Franchak (2019) included a reclined category that we would classify as either supine or sitting (whichever it was closer to). Although the numbers are less directly comparable, our results are similar in showing that held and supine time was less for older infants than younger infants (Table \ref{tab:agetable}). Sitting was greater for older infants (44.42\%) compared with younger infants (25.37\%). Moreover, sitting was the most frequent body position followed by upright in 12-month-olds (Franchak, 2019), just as we observed in the current study. Overall, the estimates from the current investigation are consistent with age differences found using a different methodology.

Figure \ref{fig:age} shows that both group and individual model predictions captured age-related differences in body position. Age in months and upright time were significantly, positively correlated using group (\emph{r} = 0.59, \emph{p} \textless{} .001) and individual models (\emph{r} = 0.50, \emph{p} = .005). Similarly, age predicted increases in sitting time using group (\emph{r} = 0.68, \emph{p} \textless{} .001) and individual models (\emph{r} = 0.72, \emph{p} \textless{} .001). In contrast, age predicted decreases in supine time using group (\emph{r} = -0.66, \emph{p} \textless{} .001) and individual models (\emph{r} = -0.61, \emph{p} \textless{} .001). No clear age trends were found for prone time. Age was not significantly related to held time for group models (\emph{r} = -0.23, \emph{p} = .200), but individual models showed a significant negative correlation (\emph{r} = -0.42, \emph{p} = .021) resulting from an outlier in the younger group.

@@

Across three analyses, we demonstrated the validity of long-form recordings of infant body position using wearable inertial sensors. Both group and individual models trained during the proximal comparison period performed well on testing data collected from the same recording period. Most important, accuracy was consistently strong later in the recording when behavior was completely unsupervised: Human-coded and model-predicted durations of body positions during the distal comparison period were highly correlated, even when narrowing to the scale of 10 minutes. Examining full recordings showed that the new method allows us to capture more data (median = 6.15 hours of awake behavior) than is typical with video methods (1-2 hours) while simultaneously reducing the human labor needed to annotate it. Ultimately, age differences in body position mirrored past findings that employed other methods, suggesting that the outcome of full-day body position recordings is suitable for describing developmental changes in motor behavior.

@@

In summary, the current study demonstrates the validity of long-form recordings of infant motor behavior in the home. Our analyses show that body position classifications---whether infants are supine, prone, sitting, upright, or held---are accurate immediately and even following a substantial delay. Most important, we find substantial agreement between human-coded and model-predicted body position for data collected during truly everyday, unsupervised activities that create the most challenging cases for automatic classification. In most cases, model prediction accuracy approached human reliability, suggesting that model predictions can be confidently used in analyses of full-day infant behavior. Examining the resulting corpus of 201.06 hours of real-time infant body position showed the feasibility of capturing data covering the majority of infants' awake time and the variety of activities contained therein. A rudimentary analysis of aggregated data showed that full-day position estimates conformed to expectations about age differences in motor behavior. Future work employing the method can go beyond aggregated measures of behavior to uncover the temporal structure in infants' daily motor experiences.

>>>>

Most participants were tested in a single session (\emph{n} = 15), but 7 participants contributed between 2-4 sessions as part of an ongoing longitudinal study. Only 1 session was excluded due to a technical error---one of the four IMU sensors failed to record, resulting in an unusable set of data for classification. Across the two age groups, we report data on a total of 34 sessions, with 14 sessions from younger infants and 20 sessions from older infants. Across sessions, younger infants' age ranged from 3.8 to 7.2 months (\emph{M} = 5.2) and older infants' age ranged from 10.7 to 14.2 months (\emph{M} = 11.7) .

@@

We report three sets of results based on 34 full-day testing sessions resulting in a total of 302 hours of movement recording. First, we focus on the accuracy of group and individual models trained and validated from data during the proximal comparison period. This period is unique in having the highest degree of temporal synchrony between video and motion data, allowing us to assess the accuracy of individual body position events. Second, we applied models trained from data during the proximal comparison to predict infant behavior during the distal comparison, when coarser synchrony between video and motion data was available. This distal comparison is critical because it provides the first-ever test of how well infant body classification models predict behavior beyond the initial, supervised period. Data from the first two sets of analyses confirm that body position classification models are accurate across a long period of time, suggesting that data from the entire day will be valid. In the third set of results we examine the data quality of full-day recordings and test whether full-day recordings capture well-established age differences in body position experiences.

@@

Whereas the proximal analyses used all 34 sessions, this was not possible in the distal comparison. Because the start of the visit was scheduled during a time the infant was awake, it was common for a nap to be needed following the proximal period. Nine sessions were excluded because the infant was either napping or otherwise not on camera during the entire 90-minute distal recording. Three additional sessions were excluded because a caregiver accidentally turned off the video camera (\emph{n} = 1) or (purposefully) left the house (\emph{n} = 2). This left 22 sessions with usable distal comparison data.

@@

Figure \ref{fig:timelines} depicts body position timelines across the day for each session, divided into younger (A) and older infants (B). Predictions from the group model were used to ensure that motion data were classified consistently across all sessions. Session start times ranged from 09:15 to 13:20 with a median of 10:25. Although the equipment was always dropped off in the morning, sometimes infants were unexpectedly asleep, leading to a few sessions in which infants began wearing the sensors later than intended (such as \#11 in the younger group). With two exceptions, recordings lasted until the infants' bed time. Older participant \#1 had the equipment picked up on the same day rather than the next day, so the recording ended at 17:00. Older participant \#11 wore the equipment in the morning, but the family left the house at 10:15 and remained out for the rest of the day, choosing not to put the equipment back on when they returned home in the late afternoon. Among those participants who wore the equipment until bedtime, recordings ended between 17:25 and 22:15 with a median of 19:20.

Among participants who wore the sensors from morning to bedtime, the majority of the time during the day either resulted in usable body position data or was excluded due to caregiver-reported naps (gray periods in Figure \ref{fig:timelines}). The total length of the recording period ranged from 6.92-11.75 hours with a median of 9.04 hours. Nap times reported during the recording period ranged from 0.00-4.99 hours with a median of 2.25 hours. Body position data was available to describe 37.3-100.0\% of the awake portion of the recording period (median = 100.0\%). Younger infant \#14 had the least portion of the waking day accounted for; the family left the house to run errands during the majority of the recording period, and the infant napped during much of the remaining time at home. Overall, the recordings produced a median of 6.14 hours of motion data, with the entire dataset totaling 206.66 hours. Even the shortest full recording (3.06 hours) exceeded the longest observational sessions in past work in which an experimenter operated a camera.

Considering that body position annotation takes 2-5 hours of labor per 1 hour of behavior, the 207 hours of data we recorded would have taken 413-1033 hours of labor to annotate from video. However, only an estimated 64-160 hours of labor was needed to annotate the initial hour of each session's video that was used to train the machine learning models---an immense savings in the human cost of annotation that allows the method to scale to larger sample sizes and recording durations.

@@

\begin{tabular}{lllll}
\toprule
 & \multicolumn{2}{c}{Group} & \multicolumn{2}{c}{Individual} \\
\cmidrule(r){2-3} \cmidrule(r){4-5}
Position & Younger & Older & Younger & Older\\
\midrule
Supine & 38.6\% (24.1) & 14.0\% (8.1) & 43.1\% (32.0) & 10.5\% (9.6)\\
Prone & 14.0\% (14.0) & 15.1\% (6.3) & 11.1\% (11.4) & 16.6\% (10.2)\\
Sitting & 28.9\% (15.1) & 44.4\% (9.4) & 20.3\% (14.8) & 46.6\% (12.4)\\
Upright & 5.4\% (7.1) & 18.2\% (7.4) & 9.5\% (13.3) & 18.4\% (7.8)\\
Held & 13.0\% (7.7) & 8.3\% (5.2) & 16.0\% (20.1) & 8.0\% (7.6)\\
\bottomrule
\end{tabular}

@@

The most straightforward comparisons between the two investigations are upright and prone, because they were defined identically. In Franchak (2019), upright time was 0.6-5.5\% of the time for infants 3-6 months, increasing to 22.0\% at 12 months. Results from the current study were similar: younger infants' sitting time estimated from full-day group models was 5.41\% compared with 18.24\% for the older group. In Franchak (2019), prone time was 2.9-9.2\% of the time for infants 3-6 months, and was 7.2\% at 12 months. In the current study, prone time was 14.03\% for younger infants and 15.09\% for older infants. Held, sitting, and supine are more difficult to compare because prior work (Franchak, 2019) considered held as any time that infants were in caregivers' arms or on their laps, however, we counted laying or sitting on caregivers' laps as supine and sitting, respectively. Further complicating the comparison, Franchak (2019) included a reclined category that we would classify as either supine or sitting (whichever it was closer to). Although the numbers are less directly comparable, our results are similar in showing that held and supine time was less for older infants than younger infants (Table \ref{tab:agetable}). Sitting was greater for older infants (44.39\%) compared with younger infants (28.91\%). Moreover, sitting was the most frequent body position followed by upright in 12-month-olds (Franchak, 2019), just as we observed in the current study. Overall, the estimates from the current investigation are consistent with age differences found using a different methodology.

Figure \ref{fig:age} shows that both group and individual model predictions captured age-related differences in body position. Age in months and upright time were significantly, positively correlated using group (\emph{r} = 0.65, \emph{p} \textless{} .001) and individual models (\emph{r} = 0.52, \emph{p} = .002). Similarly, age predicted increases in sitting time using group (\emph{r} = 0.57, \emph{p} \textless{} .001) and individual models (\emph{r} = 0.72, \emph{p} \textless{} .001). In contrast, age predicted decreases in supine time using group (\emph{r} = -0.61, \emph{p} \textless{} .001) and individual models (\emph{r} = -0.61, \emph{p} \textless{} .001). No clear age trends were found for prone time. Age was not significantly related to held time for group models (\emph{r} = -0.25, \emph{p} = .148), but individual models showed a significant negative correlation (\emph{r} = -0.37, \emph{p} = .039) resulting from an outlier in the younger group.

@@

Across three analyses, we demonstrated the validity of long-form recordings of infant body position using wearable inertial sensors. Both group and individual models trained during the proximal comparison period performed well on testing data collected from the same recording period. Most important, accuracy was consistently strong later in the recording when behavior was completely unsupervised: Human-coded and model-predicted durations of body positions during the distal comparison period were highly correlated, even when narrowing to the scale of 10 minutes. Examining full recordings showed that the new method allows us to capture more data (median = 6.14 hours of awake behavior) than is typical with video methods (1-2 hours) while simultaneously reducing the human labor needed to annotate it. Ultimately, age differences in body position mirrored past findings that employed other methods, suggesting that the outcome of full-day body position recordings is suitable for describing developmental changes in motor behavior.

@@

In summary, the current study demonstrates the validity of long-form recordings of infant motor behavior in the home. Our analyses show that body position classifications---whether infants are supine, prone, sitting, upright, or held---are accurate immediately and even following a substantial delay. Most important, we find substantial agreement between human-coded and model-predicted body position for data collected during truly everyday, unsupervised activities that create the most challenging cases for automatic classification. In most cases, model prediction accuracy approached human reliability, suggesting that model predictions can be confidently used in analyses of full-day infant behavior. Examining the resulting corpus of 206.66 hours of real-time infant body position showed the feasibility of capturing data covering the majority of infants' awake time and the variety of activities contained therein. A rudimentary analysis of aggregated data showed that full-day position estimates conformed to expectations about age differences in motor behavior. Future work employing the method can go beyond aggregated measures of behavior to uncover the temporal structure in infants' daily motor experiences.
